One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2015-01-02 paper/paper.tex
major restructuring and rewrite in response to report

\maketitle
\begin{abstract} 
  {\bf We describe a new mechanism that leads to the destabilisation of
  one-armed spiral or eccentric modes in astrophysical discs with 
  an imposed radial temperature gradient.  We use linear density wave
  theory 
  to show that non-axisymmetric perturbations generally do not conserve
  their angular momentum in the presence of a forced temperature
  gradient. This implies an exchange of angular momentum
  between linear perturbations and the background disc. In particular, 
  when the disturbance is a low-frequency, one-armed
  trailing spiral and the disc temperature decreases outwards, this 
  interaction is unstable and leads to the growth of the spiral mode.  
  We demonstrate this phenomenon through numerical hydrodynamic
  simulations of locally isothermal discs in 2D using the FARGO
  code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
  radially structured discs with a self-gravitating sub-portion which remains
  stable in the absence of a temperature gradient. However,
  when a temperature gradient is imposed we observe exponential growth of an one-armed  
  spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a low-frequency  
  spiral pattern in the self-gravitating region of the disc.  The development of this one-armed 
 spiral does not require the movement of the central star, as found in previous studies. 
 Because destabilisation by the imposed temperature gradient does 
 not explicitly depend on self-gravity, we suggest this mechanism may also
 affect low-frequency one-armed oscillations in non-self-gravitating
 discs. 
}
\end{abstract}

@@

% when/where does it happen? formation stage/dead zones (special for ppd)
GI can be expected in the earliest stages of 
circumstellar disc formation
\citep{kratter10b,inutsuka10,tsukamoto13}, and may be possible in the
outer parts of the disc \citep{rafikov05,matzner05,kimura12}.  
% outer parts  
In the case of protoplanetary discs, which have complex structure
\citep{armitage10}, GI can also occur in `dead
zones' --- sub-portions of the disc where the magneto-rotational
instability is ineffective for angular momentum transport
\citep{gammie96,turner08,landry13} and therefore has a reduced mass
accretion rate. Dead zones may then become
self-gravitating with sufficient mass built-up
\citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.

% original motivation for this work (DZGI) 
A wide range of GIs can occur in  gaseous 
astrophysical discs 
\citep{papaloizou89,christo92,christo93,hadley11,hadley14}. It is then
of theoretical importance to clarify what type of GI is relevant to 
partially self-gravitating discs, such as a protoplanetary disc containing a 
massive dead zone.  %This was indeed the original goal of this project. 
This work presents a first step towards this goal. Specifically, here we 
study a class of seemingly simple disc models which were found to 
consistently produce \emph{one-armed} spiral structures, despite not 
satisfying the usual requirements to generate them.

>>>>

\maketitle
\begin{abstract} 
  {\bf We describe a new mechanism that leads to the destabilisation of
 non-axisymmetric waves in astrophysical discs with 
  an imposed radial temperature gradient.  We use linear density wave
  theory 
  to show that non-axisymmetric perturbations generally do not conserve
  their angular momentum in the presence of a forced temperature
  gradient. This implies an exchange of angular momentum
  between linear perturbations and the background disc. In particular, 
  when the disturbance is a low-frequency
  trailing wave and the disc temperature decreases outwards, this 
  interaction is unstable and leads to the growth of the wave.  
  We demonstrate this phenomenon through numerical hydrodynamic
  simulations of locally isothermal discs in 2D using the FARGO
  code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
  radially structured discs with a self-gravitating sub-portion which remains
  stable in the absence of a temperature gradient. However,
  when a temperature gradient is imposed we observe exponential growth of an one-armed  
  spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly-stationary
  one-armed spiral pattern.  The development of this one-armed 
 spiral does not require the movement of the central star, as found in previous studies. 
 Because destabilisation by the imposed temperature gradient does 
 not explicitly depend on self-gravity, we suggest this mechanism may also
 affect low-frequency one-armed oscillations in non-self-gravitating
 discs. 
}
\end{abstract}

@@

% when/where does it happen? formation stage/dead zones (special for ppd)
GI can be expected in the earliest stages of 
circumstellar disc formation
\citep{kratter10b,inutsuka10,tsukamoto13}, and may be possible in the
outer parts of the disc \citep{rafikov05,matzner05,kimura12}.

% In the case of protoplanetary discs, which have complex structure
% \citep{armitage10}, GI can also occur in `dead
% zones' --- sub-portions of the disc where the magneto-rotational
% instability is ineffective for angular momentum transport
% \citep{gammie96,turner08,landry13} and therefore has a reduced mass
% accretion rate. Dead zones may then become
% self-gravitating with sufficient mass built-up
% \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.

% original motivation for this work (DZGI)

%A wide range of GIs can occur in  gaseous 
%astrophysical discs 
%\citep{papaloizou89,christo92,christo93,hadley11,hadley14}.

% It is then
% of theoretical importance to clarify what type of GI is relevant to 
% partially self-gravitating discs, such as a protoplanetary disc containing a 
% massive dead zone. 
% This work presents a first step towards this goal. Specifically, here we 
% study a class of seemingly simple disc models which were found to 
% consistently produce \emph{one-armed} spiral structures, despite not 
% satisfying the usual requirements to generate them.

### 2015-01-02 paper/results.tex
major restructuring and rewrite in response to report

\subsection{Reference run}
To obtain a picture of the overall disc evolution, we describe a 
fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
Fig. \ref{fargo_modeamp} plots evolution of the maximum 
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibly due to non-linear interaction of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

@@

Fig. \ref{2d_angmom} shows the evolution of disc angular momentum
components. Only the $m=0,\,1$ components are 
plotted since they are dominant. The $m=1$ structure has
an associated negative angular momentum.  
Its growth is compensated by an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 + \Delta
J_1 \sim 0$. Note that FARGO does not conserve angular momentum
exactly. However, we find the total angular momentum varies by 
$|\Delta J/J|= O(10^{-6})$, and is much smaller than the
change in the angular momenta components, $|\Delta J_{0,1}/J|>
O(10^{-5})$. Fig. \ref{2d_angmom} then suggest that angular momentum
is transferred from the one-armed spiral to the background disc.

@@

\subsection{Properties of the $m=1$ spiral and its growth}\label{fargo_m1}
In order to characterise the $m=1$ spiral that eventually
dominates, here we examine a simulation initialised with 
$M=1$ in Eq. \ref{randpert}. The one-armed spiral that emerges is more
coherent than the above simulation, which facilitates the analysis
below.

Fig. \ref{2d_fargo_viz} shows a snapshot of the $m=1$ surface
density. By measuring the $m=1$ surface density amplitude and its
pattern speed, we obtain a co-rotation radius and growth rate 
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
\end{align*}
This one-armed spiral can be considered as low frequency because its 
pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
pattern appears nearly stationary. The growth rate $\gamma$ is also slow
relative to the local rotation, although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$ is not very long.  
\begin{figure}
  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo}
  \caption{Cartesian visualisation of the $m=1$ surface density
    structure in the FARGO simulation initialised with only $m=1$
    perturbations. 
    \label{2d_fargo_viz}} 
\end{figure}

Next, we write $\Sigma_1 = 
|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
main assumption in local theory. We calculate $k$ numerically and plot
its normalised value in Fig. \ref{fargo_wavenumber}. We find 
\begin{align*}
  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
\end{align*}
where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
can apply results from local theory
self-consistently. Fig. \ref{2d_fargo_viz} shows the $m=1$ spiral is
trailing, which is consistent with $k>0$.

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. %Note that, with the above measured wavenumber $k$, 
%Eq. \ref{dispersion} requires $\nu^2 - 1 + Q^{-2}\sim0$ and the lower
%sign (long waves) to be taken. 
Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$; the bounded region is indeed 
where the $m=1$ spiral develops. This suggests that the one-armed 
spiral is trapped. Note in this region, $\nu^2 - 1 + 
Q^{-2}\simeq 0.1\ll 1$, which is necessary for consistency with 
the measured wavenumber $k$ and Eq. \ref{wavenumber}.  
%Eq. \ref{dispersion} with the lower sign taken (i.e. long waves).  
There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, acoustic waves may be launched in $R\gtrsim 7.2R_0$ by
the spiral disturbance in the inner disc \citep{lin11b}. 
%implications: effect on dynamics in outer disc

\subsubsection{Angular momentum exchange with the background disc}  
The wavenumber $k = \pi G\Sigma/c_s^2$ minimises the shifted frequency
$\sbar$ in the dispersion relation, Eq. \ref{dispersion}.  
Physically, then, it is not surprising to 
find perturbations of this wavenumber in a
self-gravitating disc where $Q\sim 1$. However, according the
local dispersion relation, $m=1$ perturbations are
formally stable for real $k$ and $Q>1$.

%most of the perturbation is inside co-rotation 
%so resonant interaction unlikely for instability

In order to track down the origin for the (slow) growth of the
$m=1$ spiral, we invoke angular momentum conservation for linear
perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
negligible at the disc boundaries, we can integrate
Eq. \ref{lin_ang_mom_cons} to give
\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_\mathrm{BG}$ is the torque density due to the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We explicitly 
compute both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
is a good match between the two torques, especially at early times
$t\lesssim110P_0$. The average discrepancy is $\simeq 5\%$. 
The match is less good later on, when the spiral
amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
less applicable. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
wave experiences a negative torque that further reduces its angular
momentum. This is consistent with angular momentum component
measurements (Fig. \ref {2d_angmom}).

% similar results if only intergrating between R_1 and R_2
%wave losses at boundaries

%ang mom decreases slightly more rapidly than torque exchange provides
%- probably non-linear effects/shocks

\subsection{Dependence on the imposed temperature profile}
The simulations above suggest the growth of the $m=1$ spiral mode
can be attributed to the background torque applied on linear perturbations
as a result of an imposed temperature gradient, as discussed in
\S\ref{global_cons}. According to Eq. \ref{baroclinic_torque}, this
torque is proportional to $q$, the power-law index used to set the
sound-speed profile (Eq. \ref{sound-speed}) in our simulations. 
We confirm this through a series of simulations with $q\in[0,1]$, 
initialised with $m=1$ perturbations. However, to keep the Toomre $Q$
profiles similar, we adjust the surface density power-law index such
that $s = (3+q)/2$.

Fig. \ref{fargo_varq} compares the $m=1$ spiral amplitudes as a
function of $q$. We indeed observe slower growth with decreasing
$q$. Although the figure indicates growth for the strictly isothermal
disc ($q=0$), we did not observe a coherent one-armed spiral upon
inspection of the $m=1$ surface density field. The growth in this case
may be associated with high-$m$ modes, which dominated the
simulation.

% difficult to measure  
% show result of isothermal calc?

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_fargo_varq}   
  \caption{Evolution of the $m=1$ spiral amplitude as a function of
    the imposed sound-speed gradient $q$. The maximum value of the
    $m=1$ surface density in $R\in[R_1,R_2]$ is shown. 
    \label{fargo_varq}} 
\end{figure}
We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a 
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
% difficult to measure very small growth rates
This shows that the temperature gradient is responsible for the
development of the one-armed spirals observed in our simulations.

@@

>>>>

\subsection{Reference run}
To obtain a picture of the overall disc evolution, we describe a 
fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
Fig. \ref{fargo_modeamp} plots evolution of the maximum 
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc {\bf is } 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibly due to non-linear interaction of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

@@

Fig. \ref{2d_angmom} shows the evolution of disc angular momentum
components. Only the $m=0,\,1$ components are 
plotted since they are dominant. The $m=1$ structure has
an associated negative angular momentum, {\bf which indicates it is a
  low-frequency mode.}   
Its growth is compensated by an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 + \Delta
J_1 \sim 0$. Note that FARGO does not conserve angular momentum
exactly. However, we find the total angular momentum varies by 
$|\Delta J/J|= O(10^{-6})$, and is much smaller than the
change in the angular momenta components, $|\Delta J_{0,1}/J|>
O(10^{-5})$. Fig. \ref{2d_angmom} then suggest that angular momentum
is transferred from the one-armed spiral to the background disc.

@@

\subsection{Dependence on the imposed temperature profile}
{\bf We show that the growth of the $m=1$ spiral is
  associated with the imposed temperature gradient  by performing a 
  series of simulations with $q\in[0,1]$,  However, to keep the Toomre $Q$
  profiles similar, we adjust the surface density power-law index such
  that $s = (3+q)/2$.  For clarity these simulations are initialised
  with $m=1$ perturbations only.  
}

% The simulations above suggest the growth of the $m=1$ spiral mode
% can be attributed to the background torque applied on linear perturbations
% as a result of an imposed temperature gradient, as discussed in
% \S\ref{global_cons}. 
% According to Eq. \ref{baroclinic_torque}, this
% torque is proportional to $q$, the power-law index used to set the
% sound-speed profile (Eq. \ref{sound-speed}) in our simulations.

Fig. \ref{fargo_varq} compares the $m=1$ spiral amplitudes as a
function of $q$. We indeed observe slower growth with decreasing
$q$. Although the figure indicates growth for the strictly isothermal 
disc ($q=0$), we did not observe a coherent one-armed spiral upon
inspection of the $m=1$ surface density field. The growth in this case
may be associated with high-$m$ modes, which dominated the 
simulation.   
% difficult to measure  
% show result of isothermal calc?
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_fargo_varq}   
  \caption{Evolution of the $m=1$ spiral amplitude as a function of
    the {\bf negative of the imposed temperature gradient} $q$. The maximum value of the
    $m=1$ surface density in $R\in[R_1,R_2]$ is shown. 
    \label{fargo_varq}} 
\end{figure}

We plot growth rates of the $m=1$ mode as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a 
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
% difficult to measure very small growth rates 
{\bf As the background torque is proportional to $q$
  (Eq. \ref{theoretical_rate})}, 
this {\bf indicates} that the temperature gradient is responsible for the
development of the one-armed spirals observed in our simulations.

@@

\subsection{Properties of the $m=1$ spiral and its growth}\label{fargo_m1}
{\bf Here we analyse the $q=1$ case in Fig. \ref{fargo_varq_growth} in
  more detail.  
}
%In order to characterise the $m=1$ spiral that eventually
%dominates, here we examine a simulation initialised with 
%$M=1$ in Eq. \ref{randpert}. The one-armed spiral that emerges is more
%coherent than the above simulation, which facilitates the analysis
%below.   
Fig. \ref{2d_fargo_viz} shows a snapshot of the $m=1$ surface
density {\bf of this run}. 
By measuring the $m=1$ surface density amplitude and its
pattern speed, we obtain a co-rotation radius and growth rate 
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
\end{align*}
This one-armed spiral can be considered as low frequency because its 
pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
pattern appears nearly stationary. The growth rate $\gamma$ is also slow
relative to the local rotation, although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$ is not very long.  
\begin{figure}
  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo}
  \caption{Cartesian visualisation of the $m=1$ surface density
    structure in the FARGO simulation initialised with only $m=1$
    perturbations. 
    \label{2d_fargo_viz}} 
\end{figure}

Next, we write $\Sigma_1 = 
|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
main assumption in local theory. We calculate $k$ numerically and plot
its normalised value in Fig. \ref{fargo_wavenumber}. We find 
\begin{align*}
  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
\end{align*}
where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
can apply {\bf results from local theory
  (\S\ref{local_approx})}. {\bf Note also that $k\simeq k_c$. }
Fig. \ref{2d_fargo_viz} shows the $m=1$
spiral is trailing, consistent with $k>0$.

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. %Note that, with the above measured wavenumber $k$, 
%Eq. \ref{dispersion} requires $\nu^2 - 1 + Q^{-2}\sim0$ and the lower
%sign (long waves) to be taken. 
Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$; the bounded region is indeed 
where the $m=1$ spiral develops. This shows that the one-armed 
spiral is trapped. Note in this region, $\nu^2 - 1 + 
Q^{-2}\simeq 0.1\ll 1$, which is necessary for consistency with 
the measured wavenumber $k$ and Eq. \ref{wavenumber}.  
%Eq. \ref{dispersion} with the lower sign taken (i.e. long waves).  
There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, acoustic waves may be launched in $R\gtrsim 7.2R_0$ by
the spiral disturbance in the inner disc \citep{lin11b}. 
%implications: effect on dynamics in outer disc

{\bf
  We can estimate the expected growth rate of the $m=1$ mode due to the
  temperature gradient. Setting $k = k_c$ and $m=1$ into
  Eq. \ref{theoretical_rate} gives
  \begin{align}\label{theoretical_rate1}
    \gamma \sim \frac{qh}{2Q}\Omega. 
  \end{align}
  Inserting $q=1$, $h=0.05$ and $Q\simeq 1.5$  gives
  $\gamma \simeq 0.017\Omega$, consistent with numerical
  results.  
}

\subsubsection{Angular momentum exchange with the background disc}  
% The wavenumber $k = \pi G\Sigma/c_s^2$ minimises the shifted frequency
% $\sbar$ in the dispersion relation, Eq. \ref{dispersion}.  
% Physically, then, it is not surprising to 
% find perturbations of this wavenumber in a
% self-gravitating disc where $Q\sim 1$. However, according the
% local dispersion relation, $m=1$ perturbations are
% formally stable for real $k$ and $Q>1$.   
%most of the perturbation is inside co-rotation 
%so resonant interaction unlikely for instability

{\bf We explicitly show that the growth of the $m=1$ spiral is due to
  the forced temperature gradient via the background torque described
  in \S\ref{wkb}. We integrate the statement for angular momentum 
  conservation for linear perturbations, Eq. \ref{lin_ang_mom_cons},
  assuming boundary fluxes are negligible, to obtain
}
% we invoke angular momentum conservation for linear
% perturbations, Eq. \ref{lin_ang_mom_cons}. 
% Assuming fluxes are
% negligible at the disc boundaries, we can integrate
% Eq. \ref{lin_ang_mom_cons} to give
\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_\mathrm{BG}$ is the torque density associated with the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We  
compute both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
is a good match between the two torques, especially at early times
$t\lesssim110P_0$. The average discrepancy is $\simeq 5\%$. 
The match is less good later on, when the spiral
amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
less applicable. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
wave experiences a negative torque that further reduces its angular
momentum, {\bf leading to its growth.} 
This is consistent with angular momentum component
measurements (Fig. \ref {2d_angmom}).

### 2015-01-02 paper/results_3d.tex
major restructuring and rewrite in response to report

In the ZEUS-MP run, we observed high-$m$ disturbances developed near 
the inner boundary initially, which is likely responsible for the
growth seen at $t<50P_0$. This is a numerical artifact and effectively
seeds the simulation with a larger perturbation. Results
from ZEUS-MP are therefore off-set from PLUTO by $\sim50P_0$. However,
once the coherent $m=1$ spiral begins to grow ($t\gtrsim 100P_0$), 
we measure similar growth rates in both codes: 
\begin{align*}
  &\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
  &\gamma \simeq 0.0085\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
\end{align*}
Both are smaller than the 2D simulations, because of the 
lower resolutions adopted in 3D and/or the small softening length
employed in 2D. 
%Rc = 8.6R_0

>>>>

In the ZEUS-MP run, we observed high-$m$ disturbances developed near 
the inner boundary initially, which is likely responsible for the
growth seen at $t<50P_0$. This is a numerical artifact and effectively
seeds the simulation with a larger perturbation. Results
from ZEUS-MP are therefore off-set from PLUTO by $\sim50P_0$. However,
once the coherent $m=1$ spiral begins to grow ($t\gtrsim 100P_0$), 
we measure similar growth rates in both codes: 
\begin{align*}
  &\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
  &\gamma \simeq 0.0085\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
\end{align*}
Both are somewhat smaller than the 2D simulations. {\bf This is
  possibily} because of the lower resolutions adopted in 3D {\bf
  and/or because the effective Toomre parameter is larger in 3D which,
  from Eq. \ref{theoretical_rate1}, is stabilising.}

% and/or the small softening length
% employed in 2D. 
%Rc = 8.6R_0
